\normalfont\spacedallcaps{Hazardous asteroids forecast via Markov random fields}

### 2021-11-26 Manuscript/manuscript.tex
PUB

In this section we are going to review the theoretical concepts on which the probabilistic methods here used are based. We will expose them following the approaches of Murphy \cite{murphy2012machine},  Koller et al. \cite{koller2009probabilistic}, H{\o}jsgaardand \cite{hojsgaard2012graphical} et al.  and Russel et al. \cite{russell2010artificial}.  Furthermore we will provide also a rapid overview of the main concepts of information theory,  following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches,  since we used some of its concepts in the preliminary analysis of the dataset. Finally we concluded this chapter with a rapid overview about the concept of algorithm intepretability, which together with the accuracy and the confusion matrix, is necessary for the assessment of a ML performances. On the other side the concepts related to the celestial mechanics here used will be described, following the Murray approach \cite{murray1999solar} into the Appendix A

@@

Lets start by supposing that we would represent compactly a joint distribution such as \cite{murphy2012machine}:

\begin{equation}
p(x_{1},x_{2},...,x_{n})
\end{equation}

that can represent for instance words in a documents or pixels of an image.  Firstly we know that using the chain rule,  we can decompose it, into the following form \cite{murphy2012machine}:

\begin{equation}
p(x_{1:V})=p(x_{1})p(x_{2}|x_{1})p(x_{3}|x_{2},x_{1})...p(x_{V}|x_{1:V-1})
\end{equation}

where V is the number of variables and 1:V stands for ${1,2,...,V}$.  This decomposition makes explicit the conditional probability tables, or in other terms the transition probability tensors \cite{wu2017markov}.  As one can point out the number of parameter is cumbersome as the number of variables grows: indeed the number of parameter required scales as $\mathcal{O}(K^{V})$. 
Such formidable problem can be attacked by considering the concept of conditional independence.  This is defined as  \cite{murphy2012machine}:

@@

\section{Information theory}
Given an ensemble of random variable, we can find the amount of information that one variable contains of another one: such quantity is called mutual information and it is a key concept within the information theory. This approach, that was implemented by Claude Shannon decades before the probabilistic modelling, represent a complementary way by which one can attack the problem of conditional dependence between random variables. Here we would provide some basic concepts of this theory, following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches, that allows to properly define the concept of mutual information. The starting concept of information theory is the entropy which express the uncertainty of a random variable. Given a random variable $X$ with alphabet (the accessible states) $\cite{cover2006elements}:$ and probability mass function $p(x)=Pr\left\lbrace X=x \rbrace\; x \in \chi$ we define the entropy of $X$ as $H(X)=-\sum_{x \in X} p(x)log p(x)$ where the logarithm has to be considered with basis 2. In analogous way the joint entropy of two random variables $(X,Y)$ with a joint distribution p(x,y) is defined as \cite{cover2006elements}:
\begin{equation}
H(X,Y)=-\sum_{x\in \chi}\sum_{y\in \mathcal{Y}}p(x,y)\log p(x,y)
\end{equation}
Furthermore, we can define also the conditional entropy as \cite{cover2006elements}:
\begin{equation}
\begin{split}
H(X|Y)=& \sum_{x\in \mathcal{X} }p(x)H(Y|X=x)\\
=& -\sum_{x\in \mathcal{X}}p(x)\sum_{y\in \mathcal{Y}}p(x,y)\log p(y|x) \\
=& -E\log p(Y|X)
\end{split}
\end{equation}
The joint entropy and the conditional entropy are related by the chain rule \cite{cover2006elements}:
\begin{equation}
H(X,Y)=H(X)+H(Y|X)
\end{equation}
Such rule can be extended to to the following from \cite{cover2006elements}:
\begin{equation}
H(X,Y|Z)=H(X|Z)+H(Y|X,Z)
\end{equation}
Given a distribution q and another distribution p, one can quantify how inefficently the second one describe the first one using the concept of relative entropy or Kullback-Leibler distance \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)=\sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}
As stated by the Gibbs inequality \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)\geq 0
\end{equation}
this quantity can not be negative: the entropy of a random variable associated to another cannot have a degree of uncertainty lower with respect to the quantity that its aimed to describe. On these basis we are now ready to introduce the concept of mutual information. This is defined as  \cite{cover2006elements}:
\begin{equation}
\begin{split}
I(X;Y)&=\sum\sump(x,y)\log\dfrac{p(x,y)}{p(x)p(y)}=\\
&=D(p(x,y)||p(x)p(y)) \\
&=H(X)-H(X|Y)=H(Y)-H(Y|X)
\end{split}
\end{equation}
As for the joint distribution also in this case we have a chain rule  \cite{cover2006elements}:
\begin{equation}
I(X_{1},X_{2},...,X_{n};Y)=\sum^{n}_{i=1}I(X_{i};Y|X_{i-1},X_{i-2},...,X_{1})
\end{equation}
Finally we would report the data process inequality theorem that connects the information theory with the Markov chain: if we have a Markov chains, $X\rightarrow Y \rightarrow Z$  then $I(X;Y)\geq I(X;Z) $. As for the Gibbs inequality, the underlying idea is that no clever manipulation of the data can improve the inference that can be made on them \cite{cover2006elements,mackay2003information}. Otherwise we would have a clear violation of the second principle of thermodynamics (see for instance the Maxwell's demon \cite{feynman2018feynman})

@@

Beside the accuracy and other features that are connected with the confusion matrix, another important feature is the algorithm intepretability. This feature, at a first instance, is connected on how much the algorithm provide explanation to a human on how it choose a forecast instead of another one, or in other word how much the algorithm is not a blackbox. Apart from all legal problems connected with an algorithm that is a blackbox\footnote{Think for instance the conditions imposed by the GDPR (e.g the right to explanation)}, in the author opinion results provided by such algorithm cannot be considered scientific, since the scientific method also require to provide an explanation and not simply almost correct forecast. Following the \cite{tarski1953undecidable} argument, which is root on mathematical logic, it can be said that the a formal theory T into S if and only if S is able to proof the theorem of T in its own language. On the other side we would also that an algorithm, as an explanation, is complete: we expect that it will be also able to make correct forecasts. However as shown by \cite{doshi2017towards} and by \cite{gilpin2018explaining}  the intepretability of an algorithm is linked with its incompleteness. This is something that is nested both in formal system as algorithms or scientific theories. For instance lets consider the fall of object: at first glance one can consider the only the items on Earth, in this case the acceleration is constant and it is a very simple theory; but as one moves to consider also the interaction within planets a much more complicate law should be considered (for which the previous case is a particular one). The first one is largely explainable but not poorly complete, the second one more difficult to explain but more complete. Therefore, in principle, the evaluation of an algorithm performances should be considered not only in to one dimension, the accuracy, but also by considering its intepretability as shown in Fig. \ref{ML_intepretability}

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into an more user-friendly mode the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research center. Among the 40 features there were present in the dataset we excluded the ones that were clearly redundant (such as the distance quantity evaluated in miles instead of kilometres), the ones that were connected only to the other name of the asteroid, the one connected to the name of the orbit and the one connected with the orbiting planet, since for all it was the Earth. Thus the features obtained were reduced to 22. Their description will be postponed to Appendix A, since a description of them cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only the 16 \% of the asteroids were hazardous, we considered to reduce the number of the non-hazardous in order to improve the portion of the hazardous ones. For this purpose we constructed a database in which the proportion hazardous/not hazardous was 1:5: thus all the hazardous one were included, while the not-hazardous were randomly selected \footnote{The author is aware that in principle the proportion should be much less unbalanced, however the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here used was a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.

>>>>

In this section we are going to review the theoretical concepts on which the probabilistic methods here used are based. We will expose them following the approaches of Murphy \cite{murphy2012machine},  Koller et al. \cite{koller2009probabilistic}, H{\o}jsgaardand \cite{hojsgaard2012graphical} et al.  and Russel et al. \cite{russell2010artificial}.  Furthermore we will provide also a rapid overview about the main concepts of information theory,  following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches,  since we used some of its concepts in the preliminary analysis of the dataset. Finally this chapter will be concluded with a rapid overview about the concept of algorithm intepretability. Such concept, together with the accuracy and the confusion matrix, is necessary for the assessment of a ML performances. On the other side the concepts related to the celestial mechanics here used will be described, following the Murray approach \cite{murray1999solar} into the Appendix A

@@

Lets start by supposing that we would represent compactly a joint distribution such as \cite{murphy2012machine}:
\begin{equation}
p(x_{1},x_{2},...,x_{n})
\end{equation}
that can represent for instance words in a documents or pixels of an image.  Firstly we know that using the chain rule,  we can decompose it, into the following form \cite{murphy2012machine}:
\begin{equation}
p(x_{1:V})=p(x_{1})p(x_{2}|x_{1})p(x_{3}|x_{2},x_{1})...p(x_{V}|x_{1:V-1})
\end{equation}
where V is the number of variables and 1:V stands for ${1,2,...,V}$.  This decomposition makes explicit the conditional probability tables, or in other terms the transition probability tensors \cite{wu2017markov}.  As one can point out, the number of parameters is cumbersome as the number of variables grows: indeed the number of parameter required scales as $\mathcal{O}(K^{V})$.  Such formidable problem can be attacked by considering the concept of conditional independence. This is defined as\cite{murphy2012machine}:

@@

\section{Information theory}
Given an ensemble of random variable, we can qunaitify the amount of information that one variable contains of another one: such quantity is called mutual information and it is a key concept within the information theory. This approach, that was implemented by Claude Shannon decades before the probabilistic modelling, represent a complementary way by which one can attack the problem of conditional dependence between random variables. Here we would provide some basic concepts of this theory, following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches, that allow to properly define the concept of mutual information. The founding concept of information theory is the entropy. This quantity express the uncertainty of a random variable. Given a random variable $X$ with alphabet (the accessible states) $\cite{cover2006elements}$ and probability mass function $p(x)=Pr\left\lbrace X=x \rbrace\; x \in \chi$ we define the entropy of $X$ as $H(X)=-\sum_{x \in X} p(x)log p(x)$ where the logarithm has to be considered with basis 2. In analogous way the joint entropy of two random variables $(X,Y)$ with a joint distribution p(x,y) is defined as \cite{cover2006elements}:
\begin{equation}
H(X,Y)=-\sum_{x\in \chi}\sum_{y\in \mathcal{Y}}p(x,y)\log p(x,y)
\end{equation}
Furthermore, we can define also the conditional entropy as \cite{cover2006elements}:
\begin{equation}
\begin{split}
H(X|Y)=& \sum_{x\in \mathcal{X} }p(x)H(Y|X=x)\\
=& -\sum_{x\in \mathcal{X}}p(x)\sum_{y\in \mathcal{Y}}p(x,y)\log p(y|x) \\
=& -E\log p(Y|X)
\end{split}
\end{equation}
The joint entropy and the conditional entropy are related by the chain rule \cite{cover2006elements}:
\begin{equation}
H(X,Y)=H(X)+H(Y|X)
\end{equation}
Such rule can be extended to to the following from \cite{cover2006elements}:
\begin{equation}
H(X,Y|Z)=H(X|Z)+H(Y|X,Z)
\end{equation}
Given a distribution q and another distribution p, one can quantify how inefficently the second one describe the first one using the concept of relative entropy or Kullback-Leibler distance \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)=\sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}
As stated by the Gibbs inequality \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)\geq 0
\end{equation}
this quantity can not be negative: the entropy of a random variable associated to another cannot have a degree of uncertainty lower with respect to the quantity that its aimed to describe. On these basis we are now ready to introduce the concept of mutual information. This is defined as  \cite{cover2006elements}:
\begin{equation}
\begin{split}
I(X;Y)&=\sum\sump(x,y)\log\dfrac{p(x,y)}{p(x)p(y)}=\\
&=D(p(x,y)||p(x)p(y)) \\
&=H(X)-H(X|Y)=H(Y)-H(Y|X)
\end{split}
\end{equation}
As for the joint distribution also in this case we have a chain rule  \cite{cover2006elements}:
\begin{equation}
I(X_{1},X_{2},...,X_{n};Y)=\sum^{n}_{i=1}I(X_{i};Y|X_{i-1},X_{i-2},...,X_{1})
\end{equation}
Finally we would report the data process inequality theorem that connects the information theory with the Markov chain: if we have a Markov chains, $X\rightarrow Y \rightarrow Z$  then $I(X;Y)\geq I(X;Z) $. As for the Gibbs inequality, the underlying idea is that no clever manipulation of the data can improve the inference that can be made on them \cite{cover2006elements,mackay2003information}. Otherwise we would have a clear violation of the second principle of thermodynamics (see for instance the Maxwell's demon \cite{feynman2018feynman})

@@

Beside the accuracy and other features that are connected with the confusion matrix, another important feature is the algorithm intepretability. This feature, at a first instance, is connected on how much the algorithm provide explanation to a human on how it choose a forecast instead of another one, or in other word how much the algorithm is not a blackbox. Apart from all legal problems connected with an algorithm that is a blackbox\footnote{Think for instance the conditions imposed by the GDPR (e.g the right to explanation)}, in the author opinion results provided by such algorithm cannot be considered scientific, since the scientific method also requires to provide an explanation and not simply almost correct forecast. Following the Tarski et al. \cite{tarski1953undecidable} argument, which is based on mathematical logic, it can be said that the a formal theory T can be translated into S if and only if S is able to proof the theorem of T in its own language. On the other side we would also that an algorithm, as an explanation, is complete: we expect that it will be also able to make correct forecasts for all available data. However as shown by \cite{doshi2017towards} and by \cite{gilpin2018explaining} the intepretability of an algorithm is linked with its incompleteness. This is something that is nested both in formal system as algorithms or scientific theories. For instance lets consider the fall of object: at first glance one can consider the only the items on Earth, in this case the acceleration is constant and it is a very simple theory; but as one moves to consider also the interaction within planets a much more complicate law should be considered (in which the previous case is a particular one). The first one is fully explainable but poorly complete, the second one more difficult to explain but much more complete. Therefore, in principle, the evaluation of an algorithm performances should be considered not only in to one dimension, the accuracy, but also by considering its intepretability as shown in Fig. \ref{ML_intepretability}

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into an more user-friendly mode the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research center. Among the 40 features that were present in the dataset we excluded the ones that were clearly redundant (such as the distance quantity evaluated in miles instead of kilometres), the ones that were connected only to the other name of the asteroid, the one connected to the name of the orbit and the one connected with the orbiting planet, since for all it was the Earth. Thus the features obtained were therefore reduced to 22. Their description will be postponed to Appendix A, since a description of them cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only the 16 \% of the asteroids were hazardous, we considered to reduce the number of the non-hazardous in order to improve the portion of the hazardous ones. For this purpose we constructed a database in which the proportion hazardous/not hazardous was 1:5: thus all the hazardous one were included, while the not-hazardous were randomly selected \footnote{The author is aware that in principle the proportion should be much less unbalanced, however the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here used was a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.
